Biodiversity temporal trends are reshaping food web structure and redundancy in riverine ecosystems

### 2025-02-07 paper/appendix.Rmd
Minor edits: sent to Loubna & Willem for review

---
title: Appendix
subtitle:
author: Alain Danet, Loubna El Madouri, Maud Mouchet, Elisa Thébault and Colin Fontaine
date: \today
output:
  bookdown::pdf_document2:
    fig_caption: true
    keep_tex: true
    toc: false
fontsize: 10pt
header-includes:
   - \usepackage{natbib}
   - \newcommand{\beginsupplement}{\renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
   - \usepackage{float}
   - \usepackage{multirow}
   - \usepackage{booktabs}
   - \usepackage{makecell}
   - \usepackage{setspace}
   - \doublespacing
   - \usepackage{lineno}
   - \linenumbers
   - \renewcommand{\arraystretch}{1}
geometry: margin=2cm
bibliography: references.bib
number_sections: false
---

@@

```{r, echo=FALSE, message=FALSE, results="hide"}
library(targets)
library(tidyverse)
library(magrittr)
library(ggeffects)
library(cowplot)
library(kableExtra)
library(sf)
library(nlme)

@@

@@

>>>>

---
title: "Biodiversity temporal trends are reshaping food web structure and redundancy in riverine ecosystems"
subtitle: "Alain Danet (a,b), Loubna El Madouri (a,c), Andrew P. Beckerman (b),
Maud Mouchet (a), Colin Fontaine (a) and Elisa Thébault (d)"
date: Supplementary Material
output:
  bookdown::pdf_document2:
    fig_caption: true
    keep_tex: true
    toc: false
fontsize: 10pt
header-includes:
   - \newcommand{\beginsupplement}{\renewcommand{\thetable}{S\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{S\arabic{figure}}}
   - \usepackage{float}
   - \usepackage{multirow}
   - \usepackage{booktabs}
   - \usepackage{makecell}
   - \usepackage{setspace}
   - \doublespacing
   - \usepackage{lineno}
   - \linenumbers
   - \renewcommand{\arraystretch}{1}
geometry: margin=2cm
bibliography: references.bib
number_sections: false
csl: pnas.csl
---

@@

```{r, echo=FALSE, message=FALSE, results="hide"}
library(targets)
library(tidyverse)
library(magrittr)
library(ggeffects)
library(cowplot)
library(kableExtra)
library(sf)
library(nlme)
library(igraph)

@@

```{r}
tar_load(c(semeff_tot, semeff_tot_ok))
semeff_tot <- semeff_tot %>%
  mutate(across(c(response, predictor), ~str_remove(.x, "mean_")))

rdtps <- map(
  setNames(
    c("connectance", "w_trph_lvl_avg", "log_bm_std", "prop_redundant_links"),
    c("ct", "tlvl", "bm", "red")
    ),
  ~ list(
    bm = get_semeff_ci(resp = .x, term = "log_bm_std"),
    r = get_semeff_ci(resp = .x, term = "log_rich_std")
    )
)
rdsp <- map(
  setNames(
    c("connectance", "w_trph_lvl_avg", "log_bm_std", "prop_redundant_links"),
    c("ct", "tlvl", "bm", "red")
    ),
  ~ list(
    bm = get_semeff_ci(resp = .x, term = "log_bm_std", model = "spatial"),
    r = get_semeff_ci(resp = .x, term = "log_rich_std", model = "spatial")
    )
)
rdm <- map(
  setNames(
    c("Connectance", "Avg trophic level", "Biomass", "Redundancy"),
    c("ct", "tlvl", "bm", "red")
    ),
  ~ list(
    bm = get_semeff_ci(x = semeff_tot_ok, type = "Direct", resp = .x, term = "Biomass", model = "Theoretical model"),
    r = get_semeff_ci(x = semeff_tot_ok, type = "Direct", resp = .x, term = "Species richness", model = "Theoretical model")
    )
)
```

```{r}
ridtps <- map(
  setNames(
    c("connectance", "w_trph_lvl_avg", "log_bm_std", "prop_redundant_links"),
    c("ct", "tlvl", "bm", "red")
    ),
  ~ list(
    r = get_semeff_ci(resp = .x, term = "log_rich_std", type = "indirect")
    )
)
ridsp <- map(
  setNames(
    c("connectance", "w_trph_lvl_avg", "log_bm_std", "prop_redundant_links"),
    c("ct", "tlvl", "bm", "red")
    ),
  ~ list(
    r = get_semeff_ci(resp = .x, term = "log_rich_std",
      model = "spatial", type = "indirect")
    )
)
ridm <- map(
  setNames(
    c("Connectance", "Avg trophic level", "Biomass", "Redundancy"),
    c("ct", "tlvl", "bm", "red")
    ),
  ~ list(
    bm = get_semeff_ci(x = semeff_tot_ok, type = "Indirect", resp = .x, term = "Biomass", model = "Theoretical model"),
    r = get_semeff_ci(x = semeff_tot_ok, type = "Indirect", resp = .x, term = "Species richness", model = "Theoretical model")
    )
)
rttps <- map(
  setNames(
    c("connectance", "w_trph_lvl_avg", "log_bm_std", "prop_redundant_links"),
    c("ct", "tlvl", "bm", "red")
    ),
  ~ list(
    bm = get_semeff_ci(resp = .x, term = "log_bm_std", type = "total"),
    r = get_semeff_ci(resp = .x, term = "log_rich_std", type = "total")
    )
)
rtsp <- map(
  setNames(
    c("connectance", "w_trph_lvl_avg", "log_bm_std", "prop_redundant_links"),
    c("ct", "tlvl", "bm", "red")
    ),
  ~ list(
    bm = get_semeff_ci(resp = .x, term = "log_bm_std",
      model = "spatial", type = "total"),
    r = get_semeff_ci(resp = .x, term = "log_rich_std",
      model = "spatial", type = "total")
    )
)
rtm <- map(
  setNames(
    c("Connectance", "Avg trophic level", "Biomass", "Redundancy"),
    c("ct", "tlvl", "bm", "red")
    ),
  ~ list(
    bm = get_semeff_ci(x = semeff_tot_ok, type = "Total", resp = .x, term = "Biomass", model = "Theoretical model"),
    r = get_semeff_ci(x = semeff_tot_ok, type = "Total", resp = .x, term = "Species richness", model = "Theoretical model")
    )
)
```

@@

# Comparison of analysis in space and time

We further assessed the consistency of the reported associations among temporal
trends by implementing two additional analyses. We first tested if the
associations among the temporal trends within sites was holding across sites,
i.e. using spatial variations in food web structure rather than temporal changes
over time (space-for-time hypothesis). Second, we performed theoretical
simulations with a bioenergetic food web model to cross-reference our site and
species specific empirical data on riverine fish with a more generalized
representation of food web dynamics known to represent accurate relationships
among our focal variables [@brose_allometric_2006; @eschenbrenner_diversity_2023;
@danet_response_2024]. With those additional empirical and theoretical data, we
assessed additional SEM models with the same structure as in Section 4.2.

As in the temporal analysis, species richness had a positive effect on community
biomass both in empirical data and in theoretical simulations (resp. $r_\delta
=$ `r rdsp$bm$r` and `r rdm$bm$r`, Fig. 3a and S3). In both
empirical data and theoretical simulations, community biomass had a positive
effect on average trophic level (resp. $r_\delta =$ `r rdsp$tlvl$bm` and
`r rdm$tlvl$bm`), and redundancy (resp. $r_\delta =$ `r rdsp$red$bm` and
`r rdm$red$bm`), as in temporal analysis. We found weak evidence of a positive
effect of spatial variation in community biomass on connectance in empirical
data ($r_\delta =$ `r rdsp$ct$bm`) but strong evidence in theoretical
simulations ($r_\delta =$ `r rdm$ct$bm`, Fig. 3a).

The effects of spatial variations in species richness on food web structure
mostly aligned with those from the temporal trend analysis, except for average
trophic level. We found a direct negative effect of species richness on
connectance both using data and model simulations (resp. $r_\delta =$
`r rdsp$ct$r` and `r rdm$ct$r`, Fig. 3a), on average trophic level
(resp. $r_\delta =$ `r rdsp$tlvl$r` and `r rdm$tlvl$r`), but a direct positive
effect on redundancy (resp. $r_\delta =$ `r rdsp$red$r` and `r rdm$red$r`).
Species richness had an indirect positive effect of species richness through
community biomass, which added up to strong total negative effects on
connectance in empirical data and simulations (resp. $r_\delta =$
`r rtsp$ct$r` and `r rtm$ct$r`) and positive on redundancy (resp. $r_\delta =$
`r rtsp$red$r` and `r rtm$red$r`). We found that
species richness had a total positive effect on average trophic level in
bioenergetic model, in agreement with temporal trends (resp. $r_\delta =$ `r rtm$tlvl$r`),
but a total negative effect in spatial variations (resp. $r_\delta =$ `r rtsp$tlvl$r`).

# Disantengling functional and redundant links

Because species biomass are involved in trophic interactions, species losses can
have cascading effects and trigger secondary extinctions [@dunne_network_2002].
In the context of biodiversity crisis, assessing the robustness of ecological
networks to species losses has been then a big question in Ecology. Previous
research showed that food-web structure, such as connectance, increased
robustness [@dunne_network_2002]. The idea is that more connected food-web will
be more stable because they contain a large number of pathways convey biomass
through the food-web. However, connectance as some limitations as a robustness
metric, because each trophic link does not contribute equally to the robustness
of an ecological network [@allesina_who_2004]. Trophic links can be categorised as
functional or redundant, depending on their role in transmitting biomass through
the food-web. Functional links are the set of links that is required to connect
every species to the basal resource of the food-web, all of the other being
redundant. Only functional links affect directly robustness, as their removal
will trigger secondary extinctions. The removal of redundant links in turn do
not directly trigger secondary extinctions, but makes the food-webs more fragile
to further extinctions [@allesina_functional_2009].

The identification of functional and redundant links is based on the use of
denominator trees [@bodini_using_2009; @allesina_who_2004;
@allesina_functional_2009]. Dominator trees allows to identify the set of
trophic links that are essential to connect all the consumers to the basal resource, i.e.
the functional links, and the redundant links, i.e. the redundant trophic pathways
from basal resource to consumers. Pioneer papers explained the algorithm to
compute the set of dominator trees and we refer to them for further
information and examples [@bodini_using_2009; @allesina_who_2004;
@allesina_functional_2009]. Here, we aim to show that while implementing this
analysis could be difficult at the time, this algorithm is now widely
available. In R, it is available in `igraph::dominator_tree()`. Below we show
how the number of functional and redundant links can be  easily computed from R
by replicating three examples presented in previous studies.

```{r, echo = TRUE}
compute_redundancy
```

Here is an example with the food-web displayed in appendix of
@bodini_using_2009:

```{r, echo = TRUE}
# Get the food-web example from Bodini et al. (2009), Appendix 1 
g <- get_bodini_graph_appendix()
# compute reduncancy
out <- compute_redundancy(g, root = "R")
as.data.frame(out)
```

```{r}
vertex_color <- "white"
vertex_size  <- 40
```

We used this function to reproduce the three examples (Fig. \@ref(fig:examples)),
which were displayed in the Fig. 1 and Appendix 1 of @bodini_using_2009, and
in the Fig. 1 of @allesina_functional_2009. The table \@ref(tab:redun) displays
the number and the proportion of functional and redundant links from those
examples.

```{r bod1}
bodinifig1 <- get_bodini_graph_fig1()

# Complete graph
p_bodinifig1_all <- ~{
  lg <- layout_as_tree(bodinifig1, root = "root")
  lg[, 2] <- -lg[, 2]
  plot(bodinifig1, layout = lg, vertex.color = vertex_color, vertex.size = vertex_size)
}

# Only functional links
bodinifig1_dtree <- dominator_tree(bodinifig1, root = "root")
p_bodinifig1_functional <- ~{
  layout <- layout_as_tree(bodinifig1_dtree$domtree, root = "root")
  layout[, 2] <- -layout[, 2]
  plot(bodinifig1_dtree$domtree, layout = layout, vertex.color = vertex_color, vertex.size = vertex_size)
}
```

```{r bod2, echo = FALSE, results = "hide"}
g <- get_bodini_graph_appendix()

# Complete graph
p_bodiniappen_all <- ~{
  lg <- layout_as_tree(g, root = "R")
  lg[, 2] <- -lg[, 2]
  plot(g, layout = lg, vertex.color = vertex_color, vertex.size = vertex_size)
}

# Only functional links
dtree <- dominator_tree(g, root = "R")
p_bodiniappen_functional <- ~{
  lg <- layout_as_tree(dtree$domtree, root = "R")
  lg[, 2] <- -lg[, 2]
  plot(
    dtree$domtree,
    layout = lg,
    vertex.color = vertex_color,
    vertex.size = vertex_size
  )
}

p_bodini <- plot_grid(
  ggdraw(p_bodinifig1_all), ggdraw(p_bodinifig1_functional),
  ggdraw(p_bodiniappen_all), ggdraw(p_bodiniappen_functional),
  labels = "auto"
)
ggsave_multiple(
  paste0("p_bodini", c(".png", ".pdf")),
  plot = p_bodini,
  path = here::here("figures"),
  scale = 2.6,
  units = c("mm"),
  width = 80,
  height = 80 * 1.5)
```

```{r all1, echo = FALSE, results = "hide"}
allesinafig1 <- get_allesina_graph(diag = FALSE)

p_allesinafig1_all <- ~{
  lg <- layout_as_tree(allesinafig1, root = "0")
  lg[, 2] <- -lg[, 2]
  plot(allesinafig1, layout = lg, vertex.color = vertex_color, vertex.size = vertex_size)
}

# Only functional links
allesinafig1_dtree <- dominator_tree(allesinafig1, root = "0")
p_allesinafig1_functional <- ~{
  layout <- layout_as_tree(allesinafig1_dtree$domtree, root = "0")
  layout[, 2] <- -layout[, 2]
  plot(
    allesinafig1_dtree$domtree,
    layout = layout,
    vertex.color = vertex_color,
    vertex.size = vertex_size
  )
}

p_examples <- plot_grid(
  ggdraw(p_bodinifig1_all), ggdraw(p_bodinifig1_functional),
  ggdraw(p_bodiniappen_all), ggdraw(p_bodiniappen_functional),
  ggdraw(p_allesinafig1_all), ggdraw(p_allesinafig1_functional),
  nrow = 3,
  labels = "auto"
)
ggsave_multiple(
  paste0("p_examples", c(".png", ".pdf")),
  plot = p_examples,
  path = here::here("figures"),
  scale = 4.0,
  units = c("mm"),
  width = 80,
  height = 80 * 1.2)
```

(ref:example-cap) Reproduced examples of categorising functional links in
food webs using `igraph::dominator_tree()` R function. Left columns (a, c, e)
displays the complete food-web while right column displays food-web with the
functional links only (b, d, f). (a,b): reproduction of Fig. 1 of
@bodini_using_2009, (c, d): reproduction of appendix figures of @bodini_using_2009,
(e, f): reproduction of fig. 1 of @allesina_functional_2009. Note that (e, f) is
a bit different from the original, as noted by the authors, they could have
removed further redundant links (i.e. $2 \to 5$, or $3 \to 5$).

```{r examples, fig.cap = "(ref:example-cap)", out.height = "80%"}
knitr::include_graphics(here::here("figures", "p_examples.png"))
```

```{r}
examples <- list(
  bodiniappendix = g,
  bodinifig1 = bodinifig1,
  allesinafig1 = allesinafig1
)
redundancy_examples <- map2_dfr(
  examples,
  list(bodiniappendix = "R", bodinifig1 = "root", allesinafig1 = "0"),
  ~compute_redundancy(.x, root = .y), .id = "reference")
```

```{r}
paper_replacement <- c(
  bodinifig1 = "Bodini et al. (2009) Fig. 1",
  bodiniappendix = "Bodini et al. (2009) Supp material 1",
  allesinafig1 = "Allesina et al. (2009) Fig. 1"
)

redundancy_var_replacement <- function() {
  c(
    reference = "References",
    functional_links = "Nb functional links",
    redundant_links = "Nb redundant links",
    total_links = "Nb links",
    prop_functional_links = "Prop. functional links",
    prop_redundant_links = "Prop. redundant links"
  )
}

redundancy_examples %>%
  select(reference, total_links,
    functional_links, redundant_links,
    prop_functional_links, prop_redundant_links
  ) %>%
  mutate(reference = paper_replacement[reference]) %>%
  rename_with(~redundancy_var_replacement()[.x]) %>%
  kable(
    digits = 2,
    row.names = FALSE,
    label = "redun",
    booktabs = T,
    caption = "Computation of the number (Nb) of functional, redundant links, and
    their proportion (Prop.) from previous published food-webs.",
    ) %>%
  collapse_rows(columns = c(1), valign = "top") %>%
  kable_styling(latex_options = "scale_down")
```

### 2025-02-07 paper/manuscript.Rmd
Minor edits: sent to Loubna & Willem for review

---
title: "Biodiversity temporal trends are reshaping food web structure and redundancy in riverine ecosystems"
subtitle: "Alain Danet (a,b), Loubna El Madouri (a,c), Andrew P. Beckerman (b),
Maud Mouchet (a), Colin Fontaine (a) and Elisa Thébault (d)"
date: "Research report"
output:
  officedown::rdocx_document:
    number_sections: false
    base_format: "bookdown::word_document2"
always_allow_html: true
number_sections: false
bibliography: references.bib
link-citations: true
#csl: ecology-letters.csl
#https://pkgs.rstudio.com/rticles/articles/examples.html
toc: false
linestretch: 2.0
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{hyperref}
  - \hypersetup{backref=true}
  - \usepackage{setspace}\doublespacing
csl: pnas.csl
---

@@

The structure of ecological networks, which describes the number, distribution,
and strength of interactions among species, is critical for understanding
species coexistence, ecosystem functioning, and stability to perturbations
[@chesson_interaction_2008; @thebault_food-web_2003; @oconnor_general_2017;
@danet_species_2021; @eschenbrenner_diversity_2023; @danet_response_2024]. Both
theory and empirical evidence demonstrate that species richness and food web
structure influence coexistence [@chesson_interaction_2008], ecosystem
productivity [@thebault_food web_2003; @wang_biodiversity_2018], temporal
stability [@danet_species_2021; @danet_response_2024;
@eschenbrenner_diversity_2023], and robustness to disturbances
[@sole_complexity_2001; @dunne_network_2002; @allesina_functional_2009;
@bodini_using_2009; @montoya_press_2009; @sanders_trophic_2018]. For instance,
keystone predators can regulate dominant competitors, preventing competitive
exclusion and maintaining species richness [@heithaus_predicting_2008;
@johnson_rarity_2007; @estes_trophic_2011]. Similarly, food webs with higher
trophic levels are associated with greater ecosystem productivity
[@wang_biodiversity_2018; @schneider_animal_2016; @wang_biodiversity_2018;
@wang_intraguild_2019; @nie_will_2023] and community stability
[@danet_species_2021; @eschenbrenner_diversity_2023; @danet_response_2024],
while the number and distribution of trophic links determine resilience to
perturbations [@dunne_network_2002; @allesina_who_2004; @bodini_using_2009;
@allesina_functional_2009; @sanders_trophic_2018]. Understanding how
biodiversity changes reshape ecological networks is therefore essential for
quantifying and mitigating anthropogenic impacts on ecosystems.

@@

To achieve these objectives, we analyzed temporal trends in species richness,
biomass, and food web structure across `r nb_op` community samplings (Fig.
\@ref(fig:matmet)a-c, Table S1). Using structural equation modeling (SEM), we
inferred the relative effects of biomass and species richness trends on food web
metrics, as well as the effect of species richness on biomass (Fig.
\@ref(fig:matmet)d). We also conducted a robustness analysis on the meta-food
web (comprising all trophic species and their interactions) to identify species
extinction scenarios that could explain observed variations in food web
structure. To test the space-for-time hypothesis, we compared temporal findings
with spatial gradients using two approaches: (i) replicating our temporal
analysis using median biodiversity and food web metrics across sites (Fig.
\@ref(fig:matmet)c & e), and (ii) simulating food web dynamics across gradients
of species richness and primary productivity using a bioenergetic model
[@brose_allometric_2006; @lajaaiti_ecologicalnetworksdynamicsjl_2025]. These
simulations allowed us to evaluate whether theoretical predictions align with
our empirical data, which are limited to fish biodiversity.

@@

Our results also reveal that food webs experiencing biomass declines undergo
significant structural changes, including the loss of top predators and reduced
connectance. These findings align with energetic and trophic biogeography
theories [@lindeman_trophic-dynamic_1942; @gravel_trophic_2011], which posit
that the presence of high-trophic-level species is constrained by the
availability of energy and resources [@lindeman_trophic-dynamic_1942,
@takimoto_ecosystem_2008; @post_ecosystem_2000; @post_long_2002]. Thus, a
decline in community biomass may lead to insufficient resources to sustain top
predators, while an increase in biomass could provide the necessary energy to
support these species [@takimoto_ecosystem_2008; @post_ecosystem_2000].
Alternatively, Biodiversity-Ecosystem Functioning theory suggests that the
presence of higher trophic levels may reduce niche overlap among producers,
enhancing productivity and community biomass [@schneider_animal_2016;
@wang_biodiversity_2018; @wang_intraguild_2019; @nie_will_2023]. The
relationship between community biomass and connectance is less clear, with a
weaker positive association compared to average trophic level. Previous
theoretical studies have reported contrasting effects of connectance on
productivity [@nie_will_2023; @wang_intraguild_2019; @thebault_food-web_2003;
@poisot_trophic_2013; @schneider_animal_2016]. For instance, while increased
connectance between herbivores and plants can either enhance or reduce
productivity [@thebault_food web_2003; @poisot_trophic_2013; @nie_will_2023],
greater connectance among carnivores generally boosts productivity by reducing
top-down control and allowing biomass to accumulate across trophic levels
[@wang_intraguild_2019; @schneider_animal_2016].

@@

Our study provides compelling evidence that biodiversity trends are profoundly
reshaping food web structure. By linking these patterns to
Biodiversity-Ecosystem Functioning and food web theory, we demonstrate that
widespread biodiversity declines likely have significant and detrimental
consequences for ecosystem functioning and stability in the face of future
perturbations. These findings suggest that assessments of food web structure,
community biomass, and species richness—which are often more feasible than
direct measurements of ecosystem functions—could serve as valuable indicators of
functional decline and metrics for evaluating the efficacy of mitigation
strategies. Importantly, our study also highlights critical avenues for future
research, stemming from its limitations. For instance, integrating food web
metrics into conservation planning will require understanding how they relate to
other facets of biodiversity, such as the presence of non-native species, and
drivers of biodiversity changes, including habitat degradation
[@danet_past_2024] and climate change [@bonnaffe_warming_2024]. Ultimately, our
work aims to bridge the gap between Biodiversity-Ecosystem Functioning and food
web theory, fostering greater integration with research on biodiversity changes.

@@

We simulated the dynamics of the `r nfw` food webs across a gradient of global carrying capacity (K,
range: `r k_range["min"]`-`r k_range["max"]`). The carrying capacity were
standardized according to the number of producers and interspecific competition
($K_i = \frac{1 + \alpha_{ij} (S - 1)}{S}K$) to prevent changes in productivity
due to system size [@ives_stability_2000; @loreau_species_2008]. We ran
simulations for 5000 timesteps. At the end of simulations, we removed dead
species and measured community biomass, species richness and food web structure.
If the final community contained disconnected producers (i.e. without consumers)
or disconnected consumers (i.e. consumers without preys), we discarded them and
run the simulations again with the same parametrization. The simulations were
done using the Julia package `EcologicalNetworksDynamics.jl`
[@lajaaiti_ecologicalnetworksdynamicsjl_2024]. Then, we computed the community
and food web metrics at the last timesteps and analyzed the data with the same
SEM described above.

>>>>

---
title: "Biodiversity temporal trends are reshaping food web structure and redundancy in riverine ecosystems"
subtitle: "Alain Danet (a,b), Loubna El Madouri (a,c), Andrew P. Beckerman (b),
Maud Mouchet (a), Colin Fontaine (a) and Elisa Thébault (d)"
date: "Research report"
output:
  officedown::rdocx_document:
    number_sections: false
    base_format: "bookdown::word_document2"
always_allow_html: true
number_sections: false
bibliography: references.bib
link-citations: true
#https://pkgs.rstudio.com/rticles/articles/examples.html
toc: false
linestretch: 2.0
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{hyperref}
  - \hypersetup{backref=true}
  - \usepackage{setspace}\doublespacing
csl: pnas.csl
---

@@

The structure of ecological networks, which describes the number, distribution,
and strength of interactions among species, is critical for understanding
species coexistence, ecosystem functioning, and stability to perturbations
[@chesson_interaction_2008; @thebault_food-web_2003; @oconnor_general_2017;
@danet_species_2021; @eschenbrenner_diversity_2023; @danet_response_2024]. Both
theory and empirical evidence demonstrate that species richness and food web
structure influence coexistence [@chesson_interaction_2008], ecosystem
productivity [@thebault_food-web_2003; @wang_biodiversity_2018], temporal
stability [@danet_species_2021; @danet_response_2024;
@eschenbrenner_diversity_2023], and robustness to disturbances
[@sole_complexity_2001; @dunne_network_2002; @allesina_functional_2009;
@bodini_using_2009; @montoya_press_2009; @sanders_trophic_2018]. For instance,
keystone predators can regulate dominant competitors, preventing competitive
exclusion and maintaining species richness [@heithaus_predicting_2008;
@johnson_rarity_2007; @estes_trophic_2011]. Similarly, food webs with higher
trophic levels are associated with greater ecosystem productivity
[@wang_biodiversity_2018; @schneider_animal_2016; @wang_biodiversity_2018;
@wang_intraguild_2019; @nie_will_2023] and community stability
[@danet_species_2021; @eschenbrenner_diversity_2023; @danet_response_2024],
while the number and distribution of trophic links determine resilience to
perturbations [@dunne_network_2002; @allesina_who_2004; @bodini_using_2009;
@allesina_functional_2009; @sanders_trophic_2018]. Understanding how
biodiversity changes reshape ecological networks is therefore essential for
quantifying and mitigating anthropogenic impacts on ecosystems.

@@

To achieve these objectives, we analyzed temporal trends in species richness,
biomass, and food web structure across `r nb_op` community samplings (Fig.
\@ref(fig:matmet)a-c, Table S1). Using structural equation modeling (SEM), we
inferred the relative effects of biomass and species richness trends on food web
metrics, as well as the effect of species richness on biomass (Fig.
\@ref(fig:matmet)d). We also conducted a robustness analysis on the meta-food
web (comprising all trophic species and their interactions) to identify species
extinction scenarios that could explain observed variations in food web
structure. To test the space-for-time hypothesis, we compared temporal findings
with spatial gradients using two approaches: (i) replicating our temporal
analysis using median biodiversity and food web metrics across sites (Fig.
\@ref(fig:matmet)c & e), and (ii) simulating food web dynamics across gradients
of species richness and primary productivity using a bioenergetic model
[@brose_allometric_2006; @lajaaiti_ecologicalnetworksdynamicsjl_nodate]. These
simulations allowed us to evaluate whether theoretical predictions align with
our empirical data, which are limited to fish biodiversity.

@@

Our results also reveal that food webs experiencing biomass declines undergo
significant structural changes, including the loss of top predators and reduced
connectance. These findings align with energetic and trophic biogeography
theories [@lindeman_trophic-dynamic_1942; @gravel_trophic_2011], which posit
that the presence of high-trophic-level species is constrained by the
availability of energy and resources [@lindeman_trophic-dynamic_1942,
@takimoto_ecosystem_2008; @post_ecosystem_2000; @post_long_2002]. Thus, a
decline in community biomass may lead to insufficient resources to sustain top
predators, while an increase in biomass could provide the necessary energy to
support these species [@takimoto_ecosystem_2008; @post_ecosystem_2000].
Alternatively, Biodiversity-Ecosystem Functioning theory suggests that the
presence of higher trophic levels may reduce niche overlap among producers,
enhancing productivity and community biomass [@schneider_animal_2016;
@wang_biodiversity_2018; @wang_intraguild_2019; @nie_will_2023]. The
relationship between community biomass and connectance is less clear, with a
weaker positive association compared to average trophic level. Previous
theoretical studies have reported contrasting effects of connectance on
productivity [@nie_will_2023; @wang_intraguild_2019; @thebault_food-web_2003;
@poisot_trophic_2013; @schneider_animal_2016]. For instance, while increased
connectance between herbivores and plants can either enhance or reduce
productivity [@thebault_food-web_2003; @poisot_trophic_2013; @nie_will_2023],
greater connectance among carnivores generally boosts productivity by reducing
top-down control and allowing biomass to accumulate across trophic levels
[@wang_intraguild_2019; @schneider_animal_2016].

@@

Our study provides compelling evidence that biodiversity trends are profoundly
reshaping food web structure. By linking these patterns to
Biodiversity-Ecosystem Functioning and food web theory, we demonstrate that
widespread biodiversity declines likely have significant and detrimental
consequences for ecosystem functioning and stability in the face of future
perturbations. These findings suggest that assessments of food web structure,
community biomass, and species richness—which are often more feasible than
direct measurements of ecosystem functions—could serve as valuable indicators of
functional decline and metrics for evaluating the efficacy of mitigation
strategies. Importantly, our study also highlights critical avenues for future
research, stemming from its limitations. For instance, integrating food web
metrics into conservation planning will require understanding how they relate to
other facets of biodiversity, such as the presence of non-native species, and
drivers of biodiversity changes, including habitat degradation
[@danet_past_2024] and climate change [@bonnaffe_interaction_2024]. Ultimately, our
work aims to bridge the gap between Biodiversity-Ecosystem Functioning and food
web theory, fostering greater integration with research on biodiversity changes.

@@

We simulated the dynamics of the `r nfw` food webs across a gradient of global carrying capacity (K,
range: `r k_range["min"]`-`r k_range["max"]`). The carrying capacity were
standardized according to the number of producers and interspecific competition
($K_i = \frac{1 + \alpha_{ij} (S - 1)}{S}K$) to prevent changes in productivity
due to system size [@ives_stability_2000; @loreau_species_2008]. We ran
simulations for 5000 timesteps. At the end of simulations, we removed dead
species and measured community biomass, species richness and food web structure.
If the final community contained disconnected producers (i.e. without consumers)
or disconnected consumers (i.e. consumers without preys), we discarded them and
run the simulations again with the same parametrization. The simulations were
done using the Julia package `EcologicalNetworksDynamics.jl`
[@lajaaiti_ecologicalnetworksdynamicsjl_nodate]. Then, we computed the community
and food web metrics at the last timesteps and analyzed the data with the same
SEM described above.
